---
title: "ESM 262 Assignment 2"
author: "Rucha Thakar"
date: "May 11, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PART 1

```{r, include=FALSE}
setwd("C:/boxsync/rthakar/Courses/Spring2017/ESM262/EnvInformatics/Assignment2")

library (tidyverse)
library (dplyr)
library (gdata)
library (lubridate)
library (pander)
```

## 1. Reading in the data as-is

```{r}
gaz_raw <- read.delim("C:/boxsync/rthakar/Courses/Spring2017/ESM262/EnvInformatics/Assignment2/CA_Features_20170401.txt", 
                      sep = "|", 
                      as.is = TRUE)
```

## 2. Selecting required columns and converting data frame to tibble

```{r}
gaz <- gaz_raw %>% select (ï..FEATURE_ID,FEATURE_NAME,
                            FEATURE_CLASS,STATE_ALPHA, COUNTY_NAME, 
                            PRIM_LAT_DEC, PRIM_LONG_DEC, SOURCE_LAT_DEC, 
                            SOURCE_LONG_DEC, ELEV_IN_M, MAP_NAME, 
                            DATE_CREATED, DATE_EDITED)

colnames(gaz) <- c("featureID","feature_name","feature_class", "state_alpha", 
                   "county_name", "primary_latitude", "primary_longitude", 
                   "source_latitude", "source_longitude", "elevation", 
                   "map_name", "date_created", "date_edited")

gaz <- as.tibble(gaz)
```

## 3. Change class to appropriate types

Using https://geonames.usgs.gov/domestic/states_fileformat.htm as reference, data types of descriptors were changed.

```{r}
gaz$primary_latitude <- as.numeric(gaz$primary_latitude)
gaz$primary_longitude <- as.numeric(gaz$primary_longitude)
gaz$source_latitude <- as.numeric(gaz$source_latitude)
gaz$source_longitude <- as.numeric(gaz$source_longitude)
gaz$elevation <- as.numeric(gaz$elevation)
gaz$date_created <- mdy (gaz$date_created)
gaz$date_edited <- mdy (gaz$date_edited)
```

## 4. Removing unknown data

According to https://geonames.usgs.gov/domestic/states_fileformat.htm, records showing "Unknown" and zeros for the latitude and longitude DMS and decimal fields, respectively, indicate that the coordinates of the feature are unknown. They are recorded in the database as zeros to satisfy the format requirements of a numerical data type. They are not errors and do not reference the actual geographic coordinates at 0 latitude, 0 longitude.

```{r}
gaz <- gaz %>% filter (primary_longitude != 0 | primary_latitude != 0)
```

In addition, drop NA values

```{r}
gaz <- gaz %>%
  drop_na(primary_latitude) %>%
  drop_na(primary_longitude)
```

## 5. Selecting only features belonging to California
Removed all features that do not belong to the state of California, USA

```{r}
California <- gaz %>% 
  filter(state_alpha == "CA")
```


## 6. Saving file to disk
Writing final file to the disk as final_file_part1.csv using "|" as separator

```{r}
write.table(California, file="California_Data.csv", sep = "|")
```


# PART 2

## 1. Most-frequently-occuring feature name in California

```{r}
Frequent_Name <- California %>% 
  count (feature_name)%>% filter (n == max(n))

Fr_Name <- Frequent_Name$feature_name[1]
Fr_Count <- Frequent_Name$n[1]
```

`r Fr_Name` is the most frequently occuring feature name in California. It occurs `r Fr_Count` times.

## 2. least-frequently-occuring feature class in California

```{r}
Rare_class <- California %>% count (feature_class)%>% filter (n == min(n))
```

`r Rare_class$feature_class[1]` and `r Rare_class$feature_class[2]` occur the least frequently in California. They occur `r Rare_class$n[1]` time and `r Rare_class$n[1]` time, respectively.

## 3. approximate center point of each county

Removing empty county names

```{r}
California <- California %>% filter (county_name != "")

County_Group <- group_by(California, county_name)
County_Mean <- summarize (County_Group, mean (primary_longitude), mean (primary_latitude))

head (County_Mean)
```

## 4. Fractions of the total number of features in each county that are natural vs. man-made

63 feature classes are categorized between natural and man-made according to the classification described in https://geonames.usgs.gov/apex/f?p=gnispq:8:0:::::

```{r}
manmade <- c("Airport", "Bridge", "Building", 
             "Canal", "Cemetery", "Census", "Church", 
             "Civil", "Crossing", "Dam", 
             "Harbor", "Hospital", "Levee", "Locale", 
             "Military", "Pillar", "Populated Place", "Post Pffice", 
             "Reservoir", "School", "Tower", "Tunnel", "Well")

California$feature <- ifelse(California$feature_class %in% manmade, "Manmade", "Natural")

Display_Data <- data.frame(California$feature_class, California$feature)
head (Display_Data)
```

Calculate fractions of natural vs man-made feature classes

```{r}
feature_count <- California %>% count (feature)

Manmade_Fraction <- feature_count$n[1]/sum(feature_count$n)
Natural_Fraction <- feature_count$n[2]/sum(feature_count$n)
```

Man-made feature classes are `r Manmade_Fraction` in proportion to all feature classes in California. Natural feature classes `r Natural_Fraction` in proportion to all feature classes in California.

